---
title: "PA1_template"
author: "SI"
date: "27 October 2017"
output: md_document

---


*Loading and preprocessing the data*

*Show any code that is needed to*

*1. Load the data (i.e. read.csv())*

```{r read data, echo=TRUE}
activity <- read.csv("activity.csv", header = TRUE)
```

*2 Process/transform the data (if necessary) into a format suitable for your analysis*

```{r transform data, echo=TRUE}
library(plyr)
library(dplyr)
library(tidyr)
library(knitr)
summary(activity)

```

*What is mean total number of steps taken per day?*

*For this part of the assignment, you can ignore the missing values in the*
*dataset.*

We take out missing values (NAs in the 'steps' variable) and assign it the a
new variable ('activityCleaned')

```{r NA cleaning, echo=TRUE}
activityCleaned <- na.omit(activity)
```

*1. Calculate the total number of steps taken per day*

After cleaning the data frame, we use the newly obtained one (activityCleaned) 
to calculate sum of steps per each day and recorded it a new Vairable
('SumStepsperDay') 

```{r Sum of steps per Day,echo=TRUE}
SumStepsperDay <- group_by(activityCleaned, date) %>% summarise(sum(steps))
SumStepsperDay
```

*2. Make a histogram of the total number of steps taken each day*

```{r Histogram Total no of Steps per Day, echo=TRUE}
hist(SumStepsperDay$`sum(steps)`, freq = TRUE, main = "Histogram for total steps per each day", xlab = "Total no of steps in a day", ylab = "Frequency (no of days with a total no of steps taken)", plot = TRUE)
```

*3. Calculate and report the mean and median of the total number of steps taken per day *

We record in a list only the sums of steps per each day ('StepsperDayOnly").
We then apply the formulas to calculate mean and median for 'StepsperDayOnly'

```{r Mean and Median of the total no of steps per day, echo=TRUE}
StepsperDayOnly <- SumStepsperDay$`sum(steps)`
mean(StepsperDayOnly)
median(StepsperDayOnly)
```

*What is the average daily activity pattern?*

*Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)*
*and the average number of steps taken, averaged across all days (y-axis)*

```{r Plot for 5 min intervals, echo=TRUE}
IntStepsMean <- activityCleaned %>% group_by(interval) %>% summarize(mean(steps))
plot(IntStepsMean, type = "l")
```


*Which 5-minute interval, on average across all the days in the dataset, *
*contains the maximum number of steps?*

We use the previously grouped  data by intervals and calculated the steps mean
by intervals in the 'IntStepsMean'.
We select and read the row with max mean, where this max appears and find 
the respective interval which is 835 (thus 8:35 to 8:40).

```{r Get interval with max mean of steps, echo=TRUE}
IntStepsMean[which.max(IntStepsMean$`mean(steps)`),]
```


*Imputing missing values*

*Note that there are a number of days/intervals where there are missing values*
*(coded as NA). The presence of missing days may introduce bias into some*
*calculations or summaries of the data.*

*1.Calculate and report the total number of missing values in the dataset (i.e.*
*the total number of rows with NAs)*


We calculate the number of NAs with the helps of is.na function, assign the
result to a variable called 'activityNAs' (in case the value is needed further)
and print this vaiable value (2304) in the same line with the help of print
function.

```{r Count NAs on steps rows, echo=TRUE}
print(activityNAs <- sum(is.na(activity$steps)))
```


*2. Devise a strategy for filling in all of the missing values in the dataset.*
*The strategy does not need to be sophisticated. For example, you could use*
*the mean/median for that day, or the mean for that 5-minute interval, etc.*

We will use the specific 5 mintue interval steps mean calculated across all
the days and replace the NA values of the steps with this mean.

*3. Create a new dataset that is equal to the original dataset but with the *
*missing data filled in.*

We have read the file into w new variable called activityUpdated and then
rewritten this variable by replacing the NA values with the mean of the steps
across all days for that 5 minute interval.

```{r New var with NAs replacing,echo=TRUE}
activityUpdated <- read.csv("activity.csv", header = TRUE)
i <- 1
repeat {
  if (is.na(activityUpdated[i, which(colnames(activityUpdated) == "steps")])) 
        {
          activityUpdated[i, which(colnames(activityUpdated) == "steps")] <- 
            filter(activityUpdated, interval == activityUpdated[i,
            which(colnames(activityUpdated) == "interval")]) %>%
              summarise(mean(steps, na.rm = TRUE))
        }
  i <- i+1
  if (i > nrow(activityUpdated)) break
}
```


*4. Make a histogram of the total number of steps taken each day and Calculate*
*and report the mean and median total number of steps taken per day. Do these*
*values differ from the estimates from the first part of the assignment? What*
*is the impact of imputing missing data on the estimates of the total daily*
*number of steps?*

```{r Histogram Total no of Steps per Day after replacing NAs, echo=TRUE}
SumStepsperDayUpdated <- group_by(activityUpdated, date) %>% summarise(sum(steps))
hist(SumStepsperDayUpdated$`sum(steps)`, freq = TRUE, main = "Histogram for total steps per each day", xlab = "Total no of steps in a day", ylab = "Frequency (no of days with a total no of steps taken)", plot = TRUE)
```

```{r Mean and Median of the total no of steps per day (NAs replaced), echo=TRUE}
StepsperDayOnly2 <- SumStepsperDayUpdated$`sum(steps)`
mean(StepsperDayOnly2)
median(StepsperDayOnly2)
```


Due to the way we did NAs replacing (means by intervals across al days), as
expected, the new mean and median are equal and also the same with the mean in
the first part: 10766.19

*Are there differences in activity patterns between weekdays and weekends?*

*For this part the weekdays() function may be of some help here. Use the*
*dataset with the filled-in missing values for this part.*


*1. Create a new factor variable in the dataset with two levels ??? ???weekday???*
*and ???weekend??? indicating whether a given date is a weekday or weekend day.*


We create a new column using mutate in order to obtain days of the week with
the help of the weekdays() function and then a separate column based on the
previous were we include weekend vs weekday labels. 


```{r Create Weekend vs Weekday labels, echo=TRUE}
activityUpdatedWEEK <- mutate(activityUpdated, DayInTheWeek = weekdays(as.Date(date)))
activityUpdatedWEEK$PartOfWeeK <- ifelse(activityUpdatedWEEK$DayInTheWeek == "Saturday" | activityUpdatedWEEK$DayInTheWeek == "Sunday", "Weekend", "Weekday")
activityUpdatedWEEK$PartOfWeeK <- factor(activityUpdatedWEEK$PartOfWeeK)
```


*2. Make a panel plot containing a time series plot (i.e. type = "l") of the*
*5-minute interval (x-axis) and the average number of steps taken, averaged*
*across all weekday days or weekend days (y-axis). See the README file in the*
*GitHub repository to see an example of what this plot should look like using*
*simulated data.*

```{r Plot, echo=TRUE}
library(lattice)
groupedByIntervalWeekDay <- ddply(activityUpdatedWEEK, ~interval + PartOfWeeK, summarise, 
    mean = mean(steps))
groupedByIntervalWeekDay$interval <- as.numeric(as.character(groupedByIntervalWeekDay$interval))
xyplot(mean ~ interval | PartOfWeeK, groupedByIntervalWeekDay, type = "l", 
    layout = c(1, 2), xlab = "Interval", ylab = "Number of steps")
```


